---
title: 'Investigation of MTCARS Data'
output:
  html_document: default
  pdf_document: default
---
Author: Pieter Overdevest

## Executive Summary
Motor Trend, a magazine about the automobile industry is interested in exploring the relationship between car variables and fuel consumption (miles per gallon). The study below shows that generally speaking manual transmission engines consume less fuel per mile than automatic transmission engines. Though, this difference seems to reduce to 0 with 8 cylinder engines. In fact, ... transmission engines will take you ... miles further per gallon than ... transmission engines. 

## Exploratory Data Analyses

First, we set a few parameters,
```{r}
library(ggplot2)
library(plyr)
library(gridExtra)

data("mtcars")
head(mtcars)
str(mtcars)
```


This shows all attributes are of class "numerical". For further assessment, we convert the levels of the transmission data to "automatic" and "manual", as this is more informative than 0 and 1 respectively. In addition, the class of the column is converted to factor,
```{r}
mtcars$am[mtcars$am == 0] <- "automatic"
mtcars$am[mtcars$am == 1] <- "manual"

mtcars$am <- as.factor(mtcars$am)
```


Next we do some exploratory data analyses on the MPG data vs transmission and the number of cylinders.
```{r}

binwidth <- (max(mtcars$mpg) - min(mtcars$mpg))/20

g <-  ggplot(data = mtcars, aes_string(x = "am", y = "mpg")) +
        geom_boxplot() +
        stat_boxplot(geom = "errorbar", width = 0.5) +
        facet_wrap(~ factor(cyl)) +
        labs(x = "Transmission", y = "Miles Per Gallon", title = "Miles per Gallon as function of the number of cylinders (4, 6 or 8) and transmission.") +
        theme(text = element_text(size = 20),
              legend.position="none") +
        geom_dotplot(binaxis = "y", stackdir = "center", stackgroups = TRUE,
              binpositions="all", binwidth = binwidth,
              dotsize = 0.6, alpha = 0.6, aes(fill = cyl))

```


## Is an automatic or manual transmission better for MPG
This will be investigated through linear modeling of the MPG data for engines with automatic vs manual transmissions. In addition, the effect of the number of cylinders will be investigated. It is anticipated that engines with a higher number cylinders will consume more gas per mile than engines with a fewer number of cylinders. 
```{r}
fit.automatic <- lm(mpg ~ cyl, data = mtcars[mtcars$am == "automatic",])
fit.manual <- lm(mpg ~ cyl, data = mtcars[mtcars$am == "manual",])

summary(fit.automatic)
summary(fit.manual)

h <- ggplot(data = mtcars, aes(x = cyl, y = mpg, col = am),
            cex=4, alpha = 0.5, width = 0.05) + 
        geom_point() +
        geom_smooth(method = "lm", se = FALSE)
h
```

Plotting the residuals shows they are reasonably around zero (),
```{r}
par(mfrow=c(1,2))
plot(mtcars$cyl[mtcars$am=="automatic"], fit.automatic$residuals, pch=19, col="red", xlab="cyl", ylab="residual / Automatic")
plot(mtcars$cyl[mtcars$am=="manual"], fit.manual$residuals, pch=19, col="blue", xlab="cyl", ylab="residual / Manual")
```

Listing the Miles Per Gallon,
```{r}
cbind(mtcars$cyl[mtcars$am=="automatic"], predict.lm(fit.automatic))
cbind(mtcars$cyl[mtcars$am=="manual"], predict.lm(fit.manual))
```
